Finiteness in Cubical Type Theory

### 2020-05-29 figures/finite-classification.tex
fix: two column figure
rearranged finite

\begin{figure}
  \centering
  \begin{tikzcd}[cramped, row sep=small]
    &
    \ar[ddddd, dash, start anchor={[yshift=4ex]}, end anchor={[yshift=-4ex]}] &
    \text{Ordered} &
    \ar[ddddd, dash, dashed, start anchor={[yshift=4ex]}, end anchor={[yshift=-4ex]}] &
    \text{Unordered}
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \ar[rrrr, dash, start anchor={[xshift=-6ex]}, end anchor={[xshift=10ex]}] &
    &
    &
    &
    {}
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    &
    &
    \text{Manifest Bishop}
      \ar[d, xshift=-1ex, bend right=30]
      \ar[rrd, end anchor=north west, start anchor=east, bend left=10, crossing over]
      &
    &
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \text{Discrete} &
    &
    \text{Split Enumerable}
      \ar[u, xshift=1ex, bend right=30]
      &
    &
    \text{Cardinal}
      \ar[ll, start anchor=west, end anchor=east, "\text{Ord}" description]
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \ar[rrrr, dash, dashed, start anchor={[xshift=-6ex]}, end anchor={[xshift=10ex]}] &
    &
    &
    &
    {}
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \text{Non Discrete} &
    {} &
    \text{Manifest Enumerable}
      \ar[from=uu, xshift=-1ex, bend right=30, crossing over]
      \ar[uu, bend right=30,  "\text{Discrete}" description]
      \ar[rr, start anchor=east, end anchor=west, crossing over]
      &
    {}
    &
    \text{Kuratowski}
      \ar[uu, xshift=1ex, bend right=30,  "\text{Discrete}" description]
      \ar[from=uu, xshift=-1ex, bend right=30, crossing over]
  \end{tikzcd}
  \caption{Classification of finiteness predicates according to whether they are
    discrete (imply decidable equality) and whether they imply a total order.}
  \label{finite-classification}
\end{figure}%

>>>>

\begin{figure*}
  \centering
  \begin{tikzcd}[cramped, row sep=small]
    {} &
    {} \ar[ddddd, dash, start anchor={[yshift=4ex]}, end anchor={[yshift=-4ex]}] &
    \text{Unordered} &
    \ar[ddddd, dash, dashed, start anchor={[yshift=4ex]}, end anchor={[yshift=-4ex]}] &
    \text{Ordered} &
    {}
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \ar[rrrr, dash, start anchor={[xshift=-6ex]}, end anchor={[xshift=10ex]}] &
    &
    &
    &
    {}
    &
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    &
    &
    &
    &
    \text{Manifest Bishop}
      \ar[d, xshift=-1ex, bend right=30]
      \ar[lld, end anchor=north east, start anchor=west, bend right=10, crossing over]
    &
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \text{Discrete} &
    &
    \text{Cardinal}
    \ar[rr, start anchor=east, end anchor=west, "\text{Ord}" description]
    &
    &
    \text{Split Enumerable}
      \ar[u, xshift=1ex, bend right=30]
    &
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \ar[rrrr, dash, dashed, start anchor={[xshift=-6ex]}, end anchor={[xshift=10ex]}] &
    &
    &
    &
    {}
    &
    \\ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \text{Non Discrete} &
    {} &
    \text{Kuratowski}
      \ar[from=uu, xshift=-1ex, bend right=30, crossing over]
      \ar[uu, bend right=30,  "\text{Discrete}" description]
      &
    {}
    &
    \text{Manifest Enumerable}
      \ar[uu, xshift=1ex, bend right=30,  "\text{Discrete}" description]
      \ar[from=uu, xshift=-1ex, bend right=30, crossing over]
      \ar[ll, start anchor=west, end anchor=east, crossing over]
    &
    \\
    & & & & &
    \\
    & \ar[rrrr, "\text{More Informative}" description, end anchor={[xshift=-6ex]}]
    &
    &
    &
    &
    \ar[uuuuuuu, "\text{More Restrictive}" description]
    {}
  \end{tikzcd}
  \caption{Classification of finiteness predicates according to whether they are
    discrete (imply decidable equality) and whether they imply a total order.}
  \label{finite-classification}
\end{figure*}%

### 2020-05-29 paper.tex
rearranged finite

Simple proof search over a finite domain is a simple, if effective, way to
relieve some of the tedium.
Many libraries in Agda, Coq, and other languages make use of this fact.
While the idea for finite proof search is simple, we found that implementing one
revealed a number of interesting theoretical considerations:
\begin{description}
  \item[How should we construct search spaces?]
    We found that a \(Pi\)-pretopos provided the appropriate ``generator
    language'' for constructing search spaces for finite types.
  \item[How do we deal with functions?]
    We of course would like to be able to search over functions: \todo{?}
  \item[How do we talk about ``searchability''?]
    \todo{Omniscience}
\end{description}
Finiteness is a prime example of a phenomenon which happens quite frequently in
dependently-typed constructive mathematics: through building a practical program
we unearth theoretically interesting results.
\todo{rephrase}
\subsection{Contributions}
We are interested in constructive notions of finiteness.
In particular, we are interested in two different aspects of each predicate for
finiteness: how informative it is (i.e. how many properties of the underlying
type we can derive from a proof that it is finite), and how restrictive it is
(i.e. how many types conform to its particular variety of finiteness).
Once we explore each of five predicates along these two axes, we will then
construct a category out of one of them (decidable Kuratowski finiteness), and
show that it forms a \(\Pi\)-pretopos.
We will then take a brief detour into predicates for countably infinite types,
and compare its landscape to that of the finite predicates, before jumping into
a practical application of the work: a library for proof search over finite
domains.
We will show how our earlier work gives a good theoretical foundation for this
library, and allows for a simple and principled API, and we will demonstrate how
some of the more complex theoretical results allow for a more powerful library
with capabilities not seen in comparable proof search libraries.

>>>>

Simple proof search over a finite domain is a simple, if effective, way to
relieve some of the tedium.
Many libraries in Agda, Coq, and other languages make use of this fact.
While the idea for finite proof search is simple, we found that implementing one
revealed a number of interesting theoretical considerations:
\begin{description}
  \item[How should we construct search spaces?]
    We found that a \(Pi\)-pretopos provided the appropriate ``generator
    language'' for constructing search spaces for finite types.
  \item[How do we deal with functions?]
    We of course would like to be able to search over functions: \todo{?}
  \item[How do we talk about ``searchability''?]
    \todo{Omniscience}
\end{description}
Finiteness is a prime example of a phenomenon which happens quite frequently in
dependently-typed constructive mathematics: through building a practical program
we unearth theoretically interesting results.
\todo{rephrase}
\subsection{Contributions}
We are interested in constructive notions of finiteness.
In contrast to classical finiteness, in a constructive setting there is a wide
variety of predicates which all have some claim to being the formal
interpretation of ``finiteness''.
Broadly, they vary along two axes: informativeness (i.e. how many properties
of the underlying type we can derive from a proof that it is finite), and
restrictiveness (i.e. how many types conform to its particular variety of
finiteness).
We will explore five of these predicates, proving formally their relations and
implications.

We will then focus on one of the predicates: decidable Kuratowski finiteness.
We will construct a category of decidable Kuratowski-finite types, and show that
it forms a \(\Pi\)-pretopos.

We will then take a brief detour into predicates for countably infinite types,
and compare its landscape to that of the finite predicates.

Finally, we will present a case-study in a practical application of the
predicates for finiteness we have constructed: a library for total proof search.
We will show how our earlier work gives a good theoretical foundation for this
library, and allows for a simple and principled API, and we will demonstrate how
some of the more complex theoretical results allow for a more powerful library
with capabilities not seen in comparable proof search libraries.
